Discrete Quantum Theories and Computing

### 2018-03-09 dissertation.tex
dissertation.lyx "Chapter 3 Quantum Theories and Computing over Finite Fields" Start to uncomment...
dissertation.lyx "Chapter 3 Quantum Theories and Computing over Finite Fields" Updating...
dissertation.lyx "2.1.2 D-dimensional Hilbert Space" Updating...

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

@@

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathbf{CP}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{DCP}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}

@@

\subsection{The single qubit problem}

@@

\subsection{The $n$-qubit problem}

For $n$ qubits, the irreducible states are encoded in a similar family
of geometric structures known technically as the complex projective
space $\CP{D-1}$. We obtain these structures starting with the $D=2^{n}$
initially unnormalized complex coefficients of the $n$-qubit state
basis 
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\ .\label{N1.eq}
\end{eqnarray}
We then follow the analog of the one-qubit procedure: Conservation
of probability requires that the norm of the vector $\VVec{\alpha}$
be normalized to unity: 
\begin{eqnarray}
\braket{\Psi}{\Psi}=\|\VVec{\alpha}\|^{2} & = & \sum_{i=0}^{D-1}|\alpha_{i}|^{2}\,=\,1\ .\label{probEq1.eq}
\end{eqnarray}
Thus the initial equation for the geometry of a quantum state describes
a \textit{topological sphere\/} $\Sphere{2D-1}$ embedded in $\mathReal^{2D}$.
To see this, remember that we can write the real and imaginary parts
of $\alpha_{i}$ as $\alpha_{i}=x_{i}+iy_{i}$, so 
\begin{eqnarray}
\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=\sum_{i=0}^{D-1}\left(x_{i}^{\ 2}+y_{i}^{\ 2}\right)=1
\end{eqnarray}
describes the locus of a $2D$-dimensional real unit vector in $\mathReal^{2D}$,
which is by definition $\Sphere{2D-1}$, the $(2D-1)$-sphere, with
$D=2^{n}$ for an $n$-qubit state.

This $\Sphere{2D-1}$ in turn is ambiguous up to the usual overall
phase, inducing an $\Sphere{1}$ symmetry action, and identifying
$\Sphere{2D-1}$ as an $\Sphere{1}$ bundle, whose base space is the
$(D-1)$-complex-dimensional projective space $\CP{D-1}$. There are
thus $2D-2$ irreducible real degrees of freedom ($D-1$ complex degrees
of freedom) for a quantum state with a $D$-dimensional basis, $\{\ket{i}\,|\,i=0,\ldots,D-1\}$.

In summary, the full space of a $D=2^{n}$-dimensional $n$-qubit
quantum state, including its overall phase defining its relationship
to other quantum states, is the topological space $\Sphere{2D-1}$.
For an isolated system, the overall phase is not measurable, and eliminating
the phase dependence in turn corresponds to identifying $\Sphere{2D-1}$
as a circle bundle over the base space $\CP{D-1}$, and therefore
$\CP{D-1}=\CP{2^{n}-1}$ defines the $2D-2$ intrinsic, irreducible,
degrees of freedom of the isolated $n$-qubit state's dynamics. In
mathematical notation, this would be written

%% From Wiki: Hopf-fibration: S^1\hookrightarrow S^3 \rightarrow S^2
\[
\Sphere{1}\hookrightarrow\Sphere{2D-1}\rightarrow\CP{D-1}\ ,
\]
with $D=2^{n}$ as usual. For $n=1$, the single qubit, we have $2^{n}-1=2-1=1$,
and the base space of the circle bundle is $\CP{1}=\Sphere{2}$, the
usual Bloch sphere. Note that only for $n=1$ is this actually a sphere-like
geometry due to an accident of low-dimensional topology.

@@

\section{Modal Quantum Computing}

\label{modalquantumcomputing}

@@

\begin{figure}[t]
\[
\begin{array}{@{\!\!}c}
\hspace*{1.35cm}\Qcircuit@C=1.3em@R=.9em{\lstick{y=\ket{0}}}\end{array}
\]
\caption[]{\label{fig:alg}Circuit for black box \usat\ in modal quantum theory
over the field $\ff{2}$. $\uf$ is a Deutsch quantum black box~\cite{NCbook}
with $\uf\ket{y}\ket{\overline{x}}~=~\ket{y\scalarPlus f(\overline{x})}\ket{\overline{x}}$,
where $\overline{x}$ denotes a sequence $x_{1},x_{2},\ldots,x_{n}$
of $n$ bits. For further notation see text.}
\end{figure}

@@

%% WE have a problem with notation - what do we use for "dot product"
%%  and/or "field norm" in the three or four instances below.
%%  Here is Yu-Tsung's comment:

% Actually, Grove didn't told about any quadratic residues, and didn't
% defined the field norm. Grove 
%  used "the quadratic form Q corresponding to a Hermitian form" instead
% of "field norm." When I wrote 
% this proof, I really read the following books:
% [12] Stewart I 2004 Galois theory (Boca Raton: Chapman and Hall/CRC).
% Field norm defined in almost all field theory books. In particular, in
% page 197 in Section 18.4  [12].
% [11] Hardy G H and Wright E M 1938 An Introduction to the Theory of
% Numbers (Oxford: Oxford University Press).
% Quadratic residues explained in almost all number theory books. In
% particular, in page 69 in Chapter VI [11].
%%  HERMITIAN (outside if needed) \define{dotprod}{dot product}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discrete Quantum Theory (I) }

\label{discretequantumtheoryI}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\emph{Discrete Quantum Computing with Complexifiable Finite Fields. }
%% NOTE (ajh): we decided Fp was complexifiable and Fpp was complexified.

\section{Discrete Quantum Computing (I)}

\label{discretequantumcomputingI}

>>>>

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[all]{xy}

@@

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}

@@

\subsection{Two-dimensional Hilbert Space}

@@

\subsection{$D$-dimensional Hilbert Space}

The irreducible states in a $D$-dimensional Hilbert space are encoded
in a similar family of geometric structures known technically as the
complex projective space $\CP{D-1}$. We obtain these structures starting
with the $D$ initially unnormalized complex coefficients of the $D$-dimensional
basis $\ket{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}\ket{i}$. We then follow
the analog of the two-dimensional procedure: Conservation of probability
requires that the norm of the vector $\VVec{\alpha}$ be normalized
to unity: 
\begin{equation}
\braket{\Psi}{\Psi}=\|\VVec{\alpha}\|^{2}=\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=1\,.\label{probEq1.eq}
\end{equation}
Thus the initial equation for the geometry of a quantum state describes
a \textit{topological sphere~}$\Sphere{2D-1}$ embedded in $\mathReal^{2D}$.
To see this, remember that we can write the real and imaginary parts
of $\alpha_{i}$ as $\alpha_{i}=x_{i}+\rmi y_{i}$, so 
\begin{equation}
\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=\sum_{i=0}^{D-1}{x_{i}}^{2}+{y_{i}}^{2}=1
\end{equation}
describes the locus of a $2D$-dimensional real unit vector in $\mathReal^{2D}$,
which is by definition $\Sphere{2D-1}$, the $(2D-1)$-sphere.

This $\Sphere{2D-1}$ in turn is ambiguous up to the usual overall
phase, inducing an $\Sphere{1}$ symmetry action, and identifying
$\Sphere{2D-1}$ as an $\Sphere{1}$ bundle, whose base space is the
$(D-1)$-complex-dimensional projective space $\CP{D-1}$. There are
thus $2D-2$ irreducible real degrees of freedom ($D-1$ complex degrees
of freedom) for a quantum state with a $D$-dimensional basis, $\set{\ket{i}}{i=0,\ldots,D-1}$.

In summary, the full space of a $D$-dimensional quantum state, including
its overall phase defining its relationship to other quantum states,
is the topological space $\Sphere{2D-1}$. For an isolated system,
the overall phase is not measurable, and eliminating the phase dependence
in turn corresponds to identifying $\Sphere{2D-1}$ as a circle bundle
over the base space $\CP{D-1}$, and therefore $\CP{D-1}$ defines
the $2D-2$ intrinsic, irreducible, degrees of freedom of the isolated
$D$-dimensional state's dynamics. In mathematical notation, this
would be written $\Sphere{1}\hookrightarrow\Sphere{2D-1}\rightarrow\CP{D-1}$.
For $D=2$, the single qubit, we have $2-1=1$, and the base space
of the circle bundle is $\CP{1}=\Sphere{2}$, the usual Bloch sphere.
Note that only for $D=2$ is this actually a sphere-like geometry
due to an accident of low-dimensional topology.

@@

\section{Modal Quantum Computing\label{modalquantumcomputing}}

@@

\begin{figure}[t]
\[
\begin{array}{@{\!\!}c}
\hspace*{1.35cm}\xymatrix@C=0pt{\lstick{y=\ket{0}} & \qw & \multigate{3}{\uf} & \gate{~S^{\dagger}~} & \ctrl{3} & \gate{~S^{\dagger}~} & \multimeasureD{3}{\text{measure}}\\
\lstick{x_{1}=\ket{0}} & \multigate{2}{\otimes S} & \ghost{\uf} & \multigate{2}{\otimes S} & \targ & \qw & \ghost{\text{measure}}\\
\lstick{\ldots} & \ghost{\otimes S} & \ghost{\uf} & \ghost{\otimes S} & \targ & \qw & \ghost{\text{measure}}\\
\lstick{x_{n}=\ket{0}} & \ghost{\otimes S} & \ghost{\uf} & \ghost{\otimes S} & \targ & \qw & \ghost{\text{measure}}
}
\end{array}
\]
\caption[]{\label{fig:alg}Circuit for black box \usat\ in modal quantum theory
over the field $\ff{2}$. $\uf$ is a Deutsch quantum black box~\cite{NCbook}
with $\uf\ket{y}\ket{\overline{x}}~=~\ket{y\scalarPlus f(\overline{x})}\ket{\overline{x}}$,
where $\overline{x}$ denotes a sequence $x_{1},x_{2},\ldots,x_{n}$
of $n$ bits. For further notation see text.}
\end{figure}

@@

Our first discrete model replaces the complex numbers by a unrestricted
finite field~$\mathbb{F}_{q}$ \cite{Schumacher2012-SCHMQT,DQT2014,SchumacherWestmoreland2010}.
In this model, an $n$-qubit state is a non-zero vector in $\mathbb{F}_{q}^{2^{n}}$.
When we measure a given state~$\ket{\phi}$, an observable is replaced
by a basis~$\mathcal{B}$ of $\mathbb{F}_{q}^{2^{n}}$. Since $\ket{\phi}$
can be represented by the summation across a unique subset~$\mathcal{S}$
of the basis~$\mathcal{B}$, whether it is possible or impossible
to measure a basis vector~$\ket{i}$ depends on whether $\ket{i}$
is in $\mathcal{S}$ or not. Because this model only predicts whether
a result is possible or impossible, it is called the modal quantum
theory. Its computational model, called the modal quantum computing,
is far from conventional quantum computing. Although modal quantum
computing can express simple algorithms such as quantum teleportation
\cite{BennettBrassardEtAl1993,peres1995quantum,Mermin2007,Jaeger2007},
it is so weak that it cannot express Deutsch's algorithm \cite{Deutsch1985,Mermin2007}.
This quantum theory is, however, also so powerful that it can be used
to solve an unstructured database search of size $N$ using $O(\log(N))$
steps, which outperforms the known asymptotic bound $O(\sqrt{N})$
in conventional quantum computing \cite{Grover:1996:FQM:237814.237866,BennettBernsteinBrassardVazirani1997,Mermin2007,Jaeger2007}.
\end{comment}

\section{Discrete Quantum Theory (I)\label{discretequantumtheoryI}}

@@

\section{Vector Spaces over Complexified Finite Fields}

In order to address the intrinsic problems induced by the notion of
the continuum calculations of the previous section, one is led to
replace the infinite fields of CQC by discrete computable fields.
Accomplishing this while maintaining the essential elements of addition
and multiplication requires a brief excursion into the theory of fields,
and particularly the theory of finite fields.

\subsection{Background}

Abstract algebra deals with various kinds of algebraic structures,
such as groups, rings, and fields, each defined by a different system
of axioms. A field~$\mathbf{F}$ is an algebraic structure consisting
of a set of elements equipped with the operations of addition, subtraction,
multiplication, and division \cite{numtheory.ref}. Fields may contain
an infinite or a finite number of elements. The rational $\mathbb{Q}$,
real $\mathbb{R}$, and complex numbers $\mathbb{C}$ are examples
of infinite fields, while the set $\mathbf{F}_{3}=\{0,1,2\}$ under
the usual multiplication and modular addition is an example of a finite
field. Finite fields are also known as Galois fields \cite{GT.ref}.

There are two distinguished elements in a field, the addition identity
$0$, and the multiplication identity $1$. Given the field $\mathbf{F}$,
the closed operations of addition, ``$+$'', and multiplication,
``$\gmult$'', satisfy the following set of axioms:
\begin{enumerate}
\item $\mathbf{F}$ is an Abelian group under the addition operation $+$
(additive group)
\item The multiplication operation $\gmult$ is associative and commutative.
The field has a multiplicative identity and the property that every
nonzero element has a multiplicative inverse
\item Distributive laws: For all $a,b,c\in\mathbf{F}$ 
\begin{eqnarray}
a\gmult(b+c)=a\gmult b+a\gmult c\\
(b+c)\gmult a=b\gmult a+c\gmult a\ .
\end{eqnarray}
\end{enumerate}
{}From now on, unless specified, we will omit the symbol $\gmult$
whenever we multiply two elements of a field.

Finite fields of $q$ elements, $\Fq=\{0,\ldots,q-1\}$, will play
a special role in this work. A simple explicit example is the following
addition and multiplication tables for $\mathbf{F}_{3}$: 
\[
\begin{array}{c|ccc}
+ & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 1 & 2\\
1 & 1 & 2 & 0\\
2 & 2 & 0 & 1
\end{array}\hspace*{2cm}\begin{array}{c|ccc}
\gmult & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 0 & 0\\
1 & 0 & 1 & 2\\
2 & 0 & 2 & 1
\end{array}
\]

\subsection{Cyclic Properties of Finite Fields}

Finite fields are classified by size. The characteristic of a field
is the least positive integer $m$ such that $m=1+1+1+\cdots+1\equiv0$,
and if no such $m$ exists we say that the field has characteristic
zero (which is the case for infinite fields such as $\mathbb{R}$).
It turns out that if the characteristic is non-zero it must be a prime
$p$. For every prime $p$ and positive integer $r$ there is a finite
field $\Fpx{p^{r}}$ of cardinality $q=p^{r}$ and characteristic
$p$ (Lagrange's theorem), which is unique up to field isomorphisms.
For every $a\in\Fq$, $a\neq0$, then $a^{q-1}=1$, implying the Frobenius
endomorphism (also a consequence of Fermat's little theorem) $a^{q}=a$,
which in turn permits us to write the multiplicative inverse of any
non-zero element in the field as $a^{-1}=a^{q-2}$, since $a^{q-2}a=a^{q-1}=1$.
Every subfield of the field $\Fq$, of cardinality $q=p^{r}$, has
$p^{r'}$ elements with some $r'$ dividing $r$, and for a given
$r'$ it is unique. Notice that a fundamental difference between finite
and infinite fields is one of topology: finite fields induce a compact
structure because of their modular arithmetic, permitting \textit{wrapping
around}, while that is not the case for fields of characteristic zero.
This feature may lead to fundamental physical consequences.

\subsection{Complexified Finite Fields}

Consider the polynomial $x^{2}+1=0$ over a finite field $\Fp$. It
is known that this polynomial does not have solutions in the field
precisely when the prime $p$ is congruent to $3\imod{4}$ (see, e.g.,
\cite{numtheory.ref}).

For such primes, it is therefore possible to construct an extended
field $\Fpp$ whose elements are of the form $\alpha=a+ib$ with $a\in\Fp$,
$b\in\Fp$, and $i$ the root of the polynomial $x^{2}+1=0$. Since
the field elements $a+ib$ behave like discrete versions of the complex
numbers, we will refer to fields $\Fp$ with prime $p$ congruent
to $3\imod{4}$ as \emph{complexifiable finite fields}, and $i$-extended
fields $\Fpp$ with $p$ congruent to $3\imod{4}$ as \emph{complexified
finite fields}.

In a complexified finite field $\Fpp$, the Frobenius automorphism
that maps $\alpha\in\Fpp$ to $\alpha^{p}\in\Fpp$ acts like complex
conjugation. For example, in $\Fppx{3}$, we have $(2+i)^{3}=8+12i-6-i=2+11i$
which, in the field, is equal to $2-i$ since $11\equiv-1\imod{3}$.

We define the \emph{field norm} $\fnorm{\cdot}$ as the map from $(a+ib)\in\Fpp$
to $a^{2}+b^{2}\in\Fp$, 
\begin{equation}
\fnorm{\alpha=a+ib}=a^{2}+b^{2}\ .\label{fieldnorm.eq}
\end{equation}
We avoid the square root in the discrete field framework because,
unlike the continuous case, the square root does not always exist.

\subsection{Vector Spaces}

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{itemize}
\item[A.] $\ip{\Phi}{\Psi}$ is the complex conjugate of $\ip{\Psi}{\Phi}$; 
\item[B.] $\ip{\Phi}{\Psi}$ is conjugate linear in its first argument and
linear in its second argument; 
\item[C.] $\ip{\Psi}{\Psi}$ is always non-negative and is equal to~0 only
if $\ket{\Psi}$ is the zero vector. 
\end{itemize}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

We are interested in the $n$-qubit vector space ${\cal H}$ of dimension
$D=2^{n}$ defined over the complexified field $\Fpp$. Let $\ket{\Psi}=(\alpha_{0}~\alpha_{1}~\ldots~\alpha_{D-1})^{T}$
and $\ket{\Phi}=(\beta_{0}~\beta_{1}~\ldots~\beta_{D-1})^{T}$ represent
vectors in ${\cal H}$, with numbers $\alpha_{i}$ and $\beta_{i}$
drawn from $\Fpp$, and where $(\cdot)^{T}$ is the transpose.

\begin{definition}{[}Hermitian dot product{]} Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, the Hermitian dot product of these vectors
is: 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}~\beta_{i}^{p}~\alpha_{i}^{\;}\ .
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions A and B for inner products but violates condition C since
in every finite field there always exists a non-zero vector $\ket{\Psi}$
such that $\ip{\Psi}{\Psi}=0$. The reason is that addition in finite
fields eventually ``wraps around'' (because of their cyclic or modular
structure), allowing the sum of non-zero elements to be zero. The
fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$ decreases
with the order $p$.

For any vector $\ket{\Psi}=(\alpha_{0}~\alpha_{1}~\ldots~\alpha_{D-1})^{T}$,
the Hermitian dot product $\ip{\Psi}{\Psi}$ is equal to $\sum_{i=0}^{D-1}~\fnorm{\alpha_{i}}$,
which is the sum of the field norms for the complex coefficients.
For convenience, we now extend the field norm to include vector arguments
by defining 
\begin{equation}
\fnorm{\ket{\Psi}}=\ip{\Psi}{\Psi}=\sum_{i=0}^{D-1}~\fnorm{\alpha_{i}}\ .\label{fnormD.eq}
\end{equation}
The field norm of a vector can vanish for non-vanishing vectors.

For vectors $\ket{\Psi}$ such that $\ip{\Psi}{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}^{p}\,\alpha_{i}^{\;}=\sum_{i=0}^{D-1}\alpha_{i}^{p+1}=\sum_{i=0}^{D-1}|\alpha_{i}|^{2}$
has a square root in the field, one can define the following ``norm'':
\begin{equation}
||\Psi||=\sqrt{\ip{\Psi}{\Psi}}\ ,\label{norm-eq}
\end{equation}
which is valid only on a subspace of the field norm for finite fields.

%There is an alternative way of defining a proper inner product over a subset of the %field $\Fpp$. Consider prime numbers $p= 8 \prod_{l=1}^k q_l -1$, where  %$q_l$ are odd primes. This subset of the primes congruent to $3 \imod 4$ has the %special properti

\section{Irreducible Discrete $n$-qubit States: Generalized Discrete Bloch
Sphere}

\label{DQCnqubitBloch.sec}

In the one-qubit state with coefficients in $\Fpp$, the discrete
analog of the Bloch sphere is constructed by exact analogy to the
continuous case: we first require that the coefficients of the single
qubit basis obey 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1\label{DQA2.eq}
\end{equation}
in the discrete field. In \ref{appU.sec}, we show that there are
$p(p^{2}-1)$ such values. Given this requirement, which is similar
in form to the conservation of probability, but not as useful due
to the lack of orderable probability values, we can immediately conclude
that the discrete analog of the Hopf fibration is again 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\label{DQA6.eq}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\ ,\nonumber 
\end{eqnarray}
but now with all computations in $(\mbox{mod}~p)$. At this point
one simply writes down all possible discrete values for the complex
numbers $\{\alpha_{0},\alpha_{1}\}$ satisfying Eq.~(\ref{DQA2.eq})
and enumerates those that project to the same value of $\{X,Y,Z\}$.
This equivalence class is the discrete analog of the circle in the
complex plane that was eliminated in the continuous case. In \ref{appH.sec},
we show that $p+1$ discrete values of $\{\alpha_{0},\alpha_{1}\}$
with unit norm map to the same point under the Hopf map \Eq{DQA6.eq};
we may think of these as discrete circles or projective lines of equivalent,
physically indistinguishable, complex phase. The surviving $p(p-1)$
values of $\{\alpha_{0},\alpha_{1}\}$ correspond to irreducible physical
states of the discrete single qubit system. %% Several examples are shown in Figure \ref{bloch.fig}. %%Thus,
for example, choosing the underlying field to be $\Fppx{3}$, there
are exactly 6 single-qubit state vectors to populate the Bloch sphere;
the four equivalent phase-multiples mapping to each of the six points
on the $\Fppx{3}$ Bloch sphere are collapsed and regarded as physically
indistinguishable. In Figure \ref{bloch.fig}, we plot the irreducible
states on the Bloch sphere for $p=3$, $7$, and $11$. Note that
the Cartesian lengths of the real vectors corresponding to the points
on the Bloch sphere vary considerably due to the nature of discrete
fields; we have artificially normalized them to a ``continuous world''
unit radius sphere for conceptual clarity.

\begin{figure}[htb]
%\vspace{-0.2in}  %% \includegraphics[width=0.9\columnwidth]{./Bloch_sphere.pdf}%%\includegraphics[width=2.0\columnwidth]{./picA.png}%%\includegraphics[width=2.0\columnwidth]{./picB.png}

\begin{centering}
\hspace*{-0.5cm}\includegraphics[width=1.1\columnwidth]{../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/picC.png}
%%\includegraphics[width=0.5\textwidth]{./picAB.eps}\vspace{-0.2in}
 %% adjust blank space?
\par\end{centering}
\caption{Schematically normalized plots of the elements of the discrete Bloch
sphere, the irreducible single-qubit (two-dimensional) state vectors
with unit norm over the field $\Fpp$. We show the results for $p=3$,
$7$, and $11$. For example, in $\Fppx{3}$, there are 24 vectors
of unit norm, but only the 6 inequivalent classes appear in the plot.
The $p+1=4$ equivalent vectors in each class differ only by a complex
discrete phase.}
\label{bloch.fig} 
\end{figure}

\subsection{Counting states on the $n$-qubit Bloch sphere}

We have the unique opportunity in the finite-field approach to quantum
computing to precisely identify and enumerate the physical states.
In the conventional theory, as we have seen, we employ a generalized
Hopf fibration on the normalized states to project out a circle of
phase-equivalent states, yielding the generalized Bloch sphere.

In the introduction to this section, we sketched the counting of the
irreducible single-qubit discrete states. To count the number of inequivalent
discrete states for the general $n$-qubit case with coefficients
in $\Fpp$, we first must find the set of unit-norm states, and then
determine the equivalence classes of unit-norm states under discrete
phase transformations; we can then enumerate the list of states on
the discrete generalized Bloch sphere. By executing computer searches
of these spaces, we discovered an hypothesis for a closed-form solution
for the counting of the states, and were then able to find a rigorous
inductive proof of the enumeration, which is presented in the Appendix.

This process of describing the discrete $n$-qubit irreducible states
can again be understood geometrically by following the discrete analog
of the Hopf fibration. First, we construct the discrete version of
the quadratic unit-length form that automatically annihilates the
distinction among states differing only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Vec{X}=(|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
{}From \ref{appH.sec}, we know that $p+1$ elements of this discrete
$\Sphere{2\times2^{n}-1}$ structure map to the \textit{same point\/}
in $\Vec{X}$. Each set of $(p+1)$ redundant points is, geometrically
speaking, the \textit{discrete Hopf fibration circle\/} living above
each \textit{irreducible\/} point of the $n$-qubit state description.
These $p+1$ points are interpretable as the $p$ finite points plus
the single point at infinity of the projective discrete line (see,
e.g., \cite{Arnold}).

The next part of this argument is the determination of the unit-norm
states, effectively the space of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The tally of unit-norm states is $p^{2^{n}-1}(p^{2^{n}}-1)$
(see \ref{appU.sec}) compared to the total number $p^{2\times2^{n}}$
of possible complex integer state vectors that could be chosen. This
unit-norm state structure is the discrete analog of $\Sphere{2\times2^{n}-1}$.

Finally, we repeat the last step of the $n$-qubit continuous Hopf
fibration process for discrete $n$-qubit states, eliminating the
discrete set of $p+1$ equivalent points that map to the same point
$\Vec{X}$ on the generalized $n$-qubit Bloch sphere. Dividing the
tally $p^{2^{n}-1}(p^{2^{n}}-1)$ of unit norm states by the $p+1$
elements of each phase-equivalent discrete circle, we find 
\[
\frac{p^{2^{n}-1}(p^{2^{n}}-1)}{p+1}=p^{2^{n}-1}(p-1)\,\prod_{k=1}^{n-1}(p^{2^{k}}+1)
\]
as the total count of unique irreducible states in a discrete $n$-qubit
configuration (see \ref{appI.sec}). The resulting object is precisely
the discrete version of $\CP{D-1}$, which we might call a \textit{discrete
complex projective space\/} or $\DCP{D-1}$, where $D=2^{n}$ as
usual.

\Comment{

\subsection{Counting states on the $n$-qubit Bloch sphere}

Although we do not have a rigorous proof for the counting of discrete
unit-norm states and the number of irreducible unit-norm states in
the discrete generalized Bloch sphere for $n$-qubit states with coefficients
in $\Fpp$, we have run extensive computer searches of the state features
for a range of compatible primes $p$ and number of qubits $n$. From
these experimental investigations, we have induced a closed-form solution
for the counting that agrees for every possible combination of parameters
that can be computed in reasonable time (the computational complexity
becomes intractable very quickly).

For $n$-qubit discrete states with coefficients in $\Fpp$, our calculation
begins with an enumeration of all possible $2^{n}$-tuples of state
coefficients in $\Fpp$; then the moduli-squared of these tuples are
summed and compared to unity in $(\mbox{mod}~p)$. Those that survive
are accumulated in a list of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The heuristic number that results for the tally of unit-norm
states is $p^{2^{n}-1}(p^{2^{n}}-1)$, compared to the total number
$p^{2\times2^{n}}$ of possible complex integer state vectors that
could be chosen. This unit-norm state structure is now the discrete
analog of $\Sphere{2\times2^{n}-1}$.
